\chapter*{Preface}

\section*{About this Book}

This book is a collection of notes and exercises based on the 2025 H2 Mathematics (9758) and H2 Further Mathematics (9649) syllabi for the Singapore-Cambridge A-Level examinations. Though this book is written primarily for students reading these courses, I warmly encourage the mathematically curious to explore its contents, especially for secondary school students that are considering reading Further Mathematics during their time in junior college.

\section*{Notation}

All definitions, resuls, methods and examples are colour-coded green, blue, purple and red respectively. 

Each result is assigned an importance rating, indicated by the number of ``$\blacktriangleright$'' symbols. The ratings are defined as follows:
\begin{itemize}
    \item {\makebox[1cm]{\hfill\difficulty{3}}} Extremely important.
    \item {\makebox[1cm]{\hfill\difficulty{2}}} Something very important.
    \item {\makebox[1cm]{\hfill\difficulty{1}}} Something important.
\end{itemize}
For exercises, particularly self-practice problems, a similar rating system is used, represented by ``\chili'' symbols.

The area of a polygon $A_1 A_2 \dots A_n$ is notated $[A_1 A_2 \dots A_n]$. In particular the area of a triangle $ABC$ is notated $[\triangle ABC]$.

For formatting reasons, an inline column vector is notated as $\cveciiix{x}{y}{z}$.

$\exp x$ is equivalent to $\e^{x}$. This is only really used when $x$ is some gnarly expression.

\section*{Contributing}

The source code for this book is available on GitHub at \href{https://github.com/asdia0/DoubleMath}{\texttt{asdia0/DoubleMath}}. Contributions are more than welcome.