\section{Self-Practice A8}

\begin{problem}
    The points $A$ and $B$ have positions vectors $\cveciiix832$ and $\cveciiix{-2}34$ respectively.

    \begin{enumerate}
        \item Show that $AB = 2\sqrt{26}$.
        \item Find the Cartesian equation for the line $AB$.
        \item The line $l$ has equation $\vec r = \cveciiix{-2}34 + t\cveciiix265$. Find the length of the projection of $AB$ onto $l$.
        \item Calculate the acute angle between $AB$ and $l$, giving your answer correct to the nearest degree.
        \item Find the position vector of the foot $N$ of the perpendicular from $A$ to $l$. Hence, find the position vector of the image of $A$ in the line $l$.
    \end{enumerate}
\end{problem}

\begin{problem}
    The position vectors of the points $A$ and $B$ are $\vec i + 2 \vec j + 3 \vec k$ and $2\vec i + 3 \vec j + p \vec k$ respectively, where $p$ is a constant. The point $C$ is such that $OABC$ is a rectangle, where $O$ is the origin.

    \begin{enumerate}
        \item Show that $p = 2$.
        \item Write down the position vector of $C$.
        \item Find a vector equation of the line $BC$.
    \end{enumerate}

    The equation of line $l$ is given by $\frac{x-1}{3} = \frac{y-1}{3}$, $z = 1$.

    \begin{enumerate}
        \setcounter{enumi}{4}
        \item Show that the lines $BC$ and $l$ are skew.
    \end{enumerate}
\end{problem}

\begin{problem}
    The lines $l_1$ and $l_2$ have equations $\vec r = \cveciiix312 + \l \cveciiix{b}1{-1}$, where $b > 1$, and $\vec r = \cveciiix401 + \m\cveciiix{-1}{-1}1$ respectively.

    \begin{enumerate}
        \item Given that the acute angle between $l_1$ and $l_2$ is $30\deg$, find the value of $b$, giving your answer correct to 3 decimal places.
    \end{enumerate}

    For the rest of the question, use $b = 3$.

    \begin{enumerate}
        \setcounter{enumi}{1}
        \item Find the coordinates of the points $A$ and $B$ where $l_1$ and $l_2$ meet the $xy$-plane respectively.
        \item The point $C$ has position vector $2\vec i + 7\vec j + 3\vec k$. Find whether $C$ is closer to $l_1$ or $l_2$.
    \end{enumerate}
\end{problem}

\begin{problem}
    Relative to an origin $O$, points $C$ and $D$ have position vectors $\cveciiix732$ and $\cveciiix{10}ab$ respectively, where $a$ and $b$ are constants.

    \begin{enumerate}
        \item The straight line through $C$ and $D$ has equation $\vec r = \cveciiix732 + t\cveciiix130$, $t \in \RR$. Find the values of $a$ and $b$.
        \item Find the position vector of the point $P$ on the line $CD$ such that $\oa{OP}$ is perpendicular to $\oa{CD}$.
        \item Find the position vector of the point $Q$ on the line $CD$ such that the angle between $\oa{OQ}$ and $\oa{OC}$ is equal to the angle between $\oa{OQ}$ and $\oa{OD}$.
    \end{enumerate}
\end{problem}

\begin{problem}
    Relative to an origin $O$, points $A$ and $B$ have position vectors $\cveciiix341$ and $\cveciiix{-1}20$ respectively. The line $l$ has vector equation $\vec r = \cveciiix6a0 + t\cveciiix13a$, where $t$ is a real parameter and $a$ is a constant. The line $m$ passes through the point $A$ and is parallel to the line $OB$.

    \begin{enumerate}
        \item Find the position vector of the point $P$ on $m$ such that $OP$ is perpendicular to $m$.
        \item Show that the two lines $l$ and $m$ have no common point.
        \item If the acute angle between the line $l$ and the $z$-axis is $60\deg$, find the exact values of the constant $a$.
    \end{enumerate}
\end{problem}

\begin{problem}
    The lines $l_1$ and $l_2$ have vector equations \[\vec r = \cveciii1{-2}3 + \l \cveciii021 \quad \land \quad \vec r = \cveciii104 + \m \cveciii1{-2}1\] respectively, where $\l$ and $\m$ are real parameters.

    \begin{enumerate}
        \item Find the acute angle between the two lines $l_1$ and $l_2$, giving your answer to the nearest $0.1\deg$.
        \item Show that $l_1$ passes through the point $P$ with position vector $\cveciiix{1}{-4}2$. Hence, show that the distance between point $P$ and any point on the line $l_2$ is given by $\sqrt{6\m^2 - 12\m + 20}$. Deduce the shortest distance between point $P$ and the line $l_2$.
    \end{enumerate}
\end{problem}

\begin{problem}[\chili]
    The coordinates of the points $A$, $B$ and $C$ are given by $A(0, 2, 4)$, $B(4, 6, 11)$ and $C(8, 1, 0)$.
    
    \begin{enumerate}
        \item Show that the triangle with vertices $A$, $B$ and $C$ is an isosceles right-angled triangle.
        \item Find the position vector of point $D$ in the same plane as $A$, $B$ and $C$ such that $BCD$ is an equilateral triangle.
    \end{enumerate}
\end{problem}

\begin{problem}[\chili]
    The equations of the lines $l_1$ and $l_2$ are given by \[l_1: \vec r = \cveciii100 + \l \cveciii123, \, \l \in \RR \quad \land \quad l_2: \vec r = \cveciii100 + \m \cveciii10{-1}, \, \m \in \RR.\]

    \begin{enumerate}
        \item The point $P$ with coordinates $(2, 2, 3)$ lies on the line $l_1$. Find the reflection of $P$ in the line $l_2$.
        \item The line $l_3$ is the reflection of the line $l_1$ in the line $l_2$. Find an equation for the line $l_3$.
        \item The line $l_4$ is such that it is parallel to $l_1$ and its distance between the two lines is $\sqrt{13/14}$. Find two possible vector equations of $l_4$.
    \end{enumerate}
\end{problem}