\section{Self-Practice A14A}

\begin{problem}
    An unbiased disc has a single dot marked on one side and two dots marked on the other side. The disc and an unbiased die are thrown, and the random variable $X$ is the sum of the number of dots showing on the disc and on the top of the die.

    \begin{enumerate}
        \item Tabulate the probability distribution of $X$.
        \item Show that $\P{X \geq 4}{X \leq 7} = 8/11$.
        \item Write down $\E{X}$ and show that $\Var{X} = 19/6$.
    \end{enumerate}
\end{problem}

\begin{problem}
    The discrete random variable $X$ denotes the number of ``sixes'' showing when two ordinary fair dice are thrown. Tabulate the probability distribution of $X$.

    Two dice are thrown repeatedly. Find the probability that, in 5 throws, the result $X=2$ occurs at least once.
    
    The dice are thrown $n$ times. Find the least value of $n$ such that \[\P{X = 2 \text{ occurs at least once in the $n$ throws}} > 0.9.\]
\end{problem}

\begin{problem}
    A writer who writes articles for a magazine finds that his proposed articles sometimes need to be revised before they are accepted for publication. The writer finds that the number of days, $X$, spent in revising a randomly chosen article can be modelled by the following discrete probability distribution.

    \begin{table}[H]
        \centering
        \begin{tabular}{|r|c|c|c|c|}
        \hline
        $x$ & 0 & 1 & 2 & 4 \\ \hline
        $\P{X = x}$ & 0.8 & 0.1 & 0.05 & 0.05 \\ \hline
        \end{tabular}
    \end{table}

    Calculate $\E{X}$ and $\Var{X}$.

    The writer prepares a series of 15 articles for the magazine. Find the expected value of the total time required for revisions to these articles.

    The writer regards articles that need no revisions (i.e. $X = 0$) or which need only minor revisions (i.e. $X = 1$) as `successful' articles, and those requiring major revisions (i.e. $X = 2$) or complete replacement (i.e. $X = 4$) as `failures'. Assuming independence, find the probability that there will be fewer than 3 `failures' in the 15 articles in the series.
\end{problem}

\begin{problem}
    In a game, 2 red balls and 8 blue balls are placed in a bottle. The bottle is shaken and Mary draws 3 balls at random without replacement. The number of red balls that the draws is denoted by $R$. Find the probability distribution of $R$, and show that $\P{R \geq 1} = 8/15$.

    Show that the expectation of $R$ is $3/5$ and find the variance of $R$.

    Mary scores 4 points for each red ball that she draws. The balls are now replaced in the bottle and the bottle is shaken again. John draws 3 balls at random and without replacement. He scores 1 point for each blue ball that he draws. Mary's score is denoted by $M$ and John's score is denoted by $J$. Find the expectation and variance of $M-J$.
\end{problem}

\begin{problem}[\chili]
    A fair cubical die has three faces marked with a `1', two faces marked with a `2' and one face marked with a `3'.

    \begin{enumerate}
        \item Calculate the expectation and variance of the score obtained when this die is thrown once.
        \item Deduce the expectation and variance of the score obtained in one throw of a second cubical die, which has one face marked `1', two faces marked `2' and three faces marked `3'.
        \item Two of the first type of die and one of the second type are thrown together, and $X$ denotes the total score obtained. Denoting the expectation and variance of $X$ by $\m$ and $\s^2$ respectively, show that $\s^2 = 5/3$ and $\P{\abs{X - \m} > 2\s} = 1/18$.
    \end{enumerate}
\end{problem}