\section{Assignment A10.1}

\begin{problem}
    The complex number $w$ is such that $w w\conj + 2w = 3 + 4\i$, where $w\conj$ is the complex conjugate of $w$. Find $w$ in the form $a + \i b$, where $a$ and $b$ are real.
\end{problem}
\begin{solution}
    Let $w = a + b \i$, where $a$ and $b$ are real. Note $w w\conj = a^2 + b^2 \in \RR$. The equation thus becomes \[a^2 + b^2 + 2\bp{a + b \i} = 3 + 4 \i.\] Comparing real and imaginary parts, we have \[a^2 + b^2 + 2a = 3 \quad \tand \quad 2b = 4.\] From the second equation, we get $b = 2$. Substituting this into the first equation, we get the quadratic $a^2 + 2a + 1 = (a+1)^2 = 0$, so $a = -1$. Thus, $w = -1 + 2\i$.
\end{solution}

\begin{problem}
    Express $(3 - \i)^2$ in the form $a + \i b$.

    Hence, or otherwise, find the roots of the equation $(z+\i)^2 = -8+6\i$.
\end{problem}
\begin{solution}
    We have $(3 - \i)^2 = 3^2 - 6 \i + \i^2 = 8 - 6\i$.

    Consider $(z+\i)^2 = -8+6\i$. Multiplying both sides by $\i^2 = -1$, we find that $(\i z - 1)^2 = 8 - 6\i$. Taking roots, we obtain $\i z - 1 = \pm (3 - \i)$, so \[z = \frac1\i (1 \pm (3 - \i)) = -\i(1 \pm (3 - \i)) = -1-4\i \tor 1 + 2\i.\]
\end{solution}

\begin{problem}
    \begin{enumerate}
        \item It is given that $z_1 = 1 + \sqrt3 \i$. Find the value of $z_1^3$, showing clearly how you obtain your answer.
        \item Given that $1 + \sqrt3 \i$ is a root of the equation \[2z^3 + az^2 + bz + 4 = 0\] find the values of the real numbers $a$ and $b$. Hence, solve the above equation.
    \end{enumerate}
\end{problem}
\begin{solution}
    \begin{ppart}
        We have \[z_1^3 = \bp{1 + \sqrt3 \i}^3 = 1 + 3\bp{\sqrt3 \i} + 3\bp{\sqrt3 \i}^2 + \bp{\sqrt3 \i}^3 = 1 + 3\sqrt 3 \i - 9 - 3\sqrt3 \i = -8.\]
    \end{ppart}
    \begin{ppart}
        Since $1 + \sqrt3\i$ is a root of the given equation, we have \[2\bp{1 + \sqrt3\i}^3 + a\bp{1 + \sqrt3\i}^2 + b\bp{1 + \sqrt3 \i} + 4 = 0.\] Expanding and simplifying, we obtain $(-2a + b) + \sqrt3 (2a + b) \i = 12$. Comparing real and imaginary parts, we obtain $-2a + b = 12$ and $2a + b = 0$, whence $a = -3$ and $b = 6$.

        Since the coefficients of $2z^3 + az^2 + bz + 4$ are all real, the second root is $(1 + \sqrt3\i)\conj = 1 - \sqrt3 \i$. Let the third root be $\a$. By Vieta's formula, \[\bp{1 + \sqrt3\i}\bp{1 - \sqrt3\i}\a = -\frac42,\] so $4\a = -2$, whence $\a = -1/2$. The roots of the equation are hence $1 + \sqrt3\i$, $1 - \sqrt3\i$ and $-1/2$.
    \end{ppart}
\end{solution}

\begin{problem}
    The complex number $z$ is such that $az^2 + bz + a = 0$ where $a$ and $b$ are real constants. It is given that $z = z_0$ is a solution to this equation where $\Im{z_0} \neq 0$.

    \begin{enumerate}
        \item Verify that $z = 1/z_0$ is the other solution. Hence, show that $\abs{z_0} = 1$.
    \end{enumerate}

    Take $\Im{z_0} = 1/2$ for the rest of the question.

    \begin{enumerate}
        \setcounter{enumi}{1}
        \item Find the possible complex numbers for $z_0$.
        \item If $\Re{z_0} > 0$, find $b$ in terms of $a$.
    \end{enumerate}
\end{problem}
\begin{solution}
    \begin{ppart}
        Observe that \[a \bp{\frac1{z_0}}^2 + b\bp{\frac1{z_0}} + a = \bp{\frac1{z_0}}^2 \bp{a + bz_0 + az_0^2} = 0.\] Hence, $z = 1/z_0$ is a root of the given equation.

        Since both $a$ and $b$ are real, by the conjugate root theorem, $z_0\conj = 1/z_0$. Hence, $\abs{z_0} = \sqrt{z_0 z_0\conj} = 1$.
    \end{ppart}
    \begin{ppart}
        Let $z_0 = x + \i/2$. Since $\abs{z_0}^2 = 1$, we have $x^2 + (1/2)^2 = 1$, so  $x^2 = 3/4$ and hence $x = \pm \sqrt3/2$. Thus, \[z_0 = \frac{\sqrt3}{2} + \frac12 \i \quad \tor \quad z_0 = -\frac{\sqrt3}{2} + \frac12 \i.\]
    \end{ppart}
    \begin{ppart}
        Since $\Re{z_0} > 0$, we have $z_0 = \sqrt{3}/2 + \i/2$. By Vieta's formula, \[-\frac{b}{a} = z_0 + \frac1{z_0} = z_0 + z_0\conj = 2 \Re{z_0} = \sqrt3,\] so $b = -\sqrt3 a$.
    \end{ppart}
\end{solution}