\section{Problem Set 5}

\begin{problem}
    Let $a$, $b$ and $n > 1$ be integers. Prove that if $m > 1$ is a divisor of $n$ and $a \equiv b \pmod{n}$, then $a \equiv b \pmod{m}$.
\end{problem}
\begin{proof}
    Since $m$ is a divisor of $n$, there exists some integer $k$ such that $n = mk$. Since $a \equiv b \pmod{n}$, there exists some integer $l$ such that \[a - b = ln = (lk)m.\] Since $lm$ is an integer, by definition, $a \equiv b \pmod{m}$.
\end{proof}

\begin{problem}
    Prove that if $3 \nmid a$, then $3 \mid a^2 + 5$.
\end{problem}
\begin{proof}
    Since $3 \nmid a$, by Fermat's Little Theorem, $a^2 +5 \equiv 1 + 5 \equiv 0 \pmod{3}$. Hence, $3 \mid a^2 + 5$.
\end{proof}

\begin{problem}
    If $m$ and $n$ are any two integers with the same parity, then $4 \mid m^2 - n^2$.
\end{problem}
\begin{proof}
    Since $m$ and $n$ have the same parity, both $m+n$ and $m-n$ are even. Hence, $m^2 - n^2 = (m+n)(m-n)$ have two factors of 2, thus it is divisible by 4.
\end{proof}

\begin{problem}
    For any integers $n$, if 3 divides $1 + n + n^2$, then $n \equiv 1 \pmod{3}$.

    \noindent\textbf{Extension.} If $p$ is a prime and $p \mid 1 + n + n^2 + \dots + n^{p-1}$, then $n \equiv 1 \pmod{p}$.
\end{problem}
\begin{proof}
    Since $3 \mid 1 + n + n^2$, we have $3 \nmid n$. Hence, by Fermat's Little Theorem, \[0 = 1 + n + n^2 \equiv 2 + n \pmod{3}.\] Thus, $n \equiv 1 \pmod{3}$.
\end{proof}
\begin{proof}[Proof of Extension.]
    We have \[0 \equiv 1 + n + n^2 + \dots + n^{p-1} = \frac{n^p - 1}{n - 1} \pmod{p}.\] Since $p$ is prime, $\ZZ / p\ZZ$ is an integral domain, so $n^p - 1 \equiv 0 \pmod{p}$. Invoking Fermat's Little Theorem, $n - 1 \equiv 0 \pmod{p}$ and the desired result immediately follows.
\end{proof}

\begin{problem}
    Prove that there is no smallest positive rational number.
\end{problem}
\begin{proof}
    Seeking a contradiction, suppose there exists a smallest positive rational number, say $x = m/n$, where $m, n \in \ZZ^+$ (WLOG) and $n \neq 0$. Consider $x/2 = m/2n$. Clearly, $x/2$ is positive and rational. Thus, $0 < x/2 < x$, contradicting the minimality of $x$. Hence, there is no smallest positive rational number.
\end{proof}

\begin{problem}
    A sequence is defined by $a_1 = 2$, $a_2 = 4$, $a_{n+2} = 5a_{n+1} - 6a_n$ for all $n \geq 1$. Prove that $a_2 = 2^n$ for all $n \in \ZZ^+$.
\end{problem}
\begin{proof}
    The characteristic equation of $a_n$ is $m^2 - 5m + 6 = (m-3)(m-2)$. Hence, \[a_n = A \cdot 2^n + B \cdot 3^n\] for some constants $A$ and $B$. From the initial conditions, \[a_1 = 2 = 2A + 3B, \quad a_2 = 4 = 4A + 9B.\] Solving, we get $A = 1$ and $B = 0$. Thus, $a_n = 2^n$.
\end{proof}

\begin{problem}
    Let $f_1, f_2, \dots$ denote the Fibonacci sequence. Show that, for all $n \in \ZZ^+$, $f_2 + f_4 + \dots + f_{2n} = f_{2n+1} - 1$.
\end{problem}
\begin{proof}
    Let $P(n)$ be the statement $f_2 + f_4 + \dots + f_{2n} = f_{2n+1} - 1$. Since $f_2 = 1 = 2 - 1 = f_3 - 1$, $P(1)$ is true. Likewise, since $f_2 + f_4 = 1 + 3 = 5 - 1 = f_5 - 1$, $P(2)$ is also true.

    Suppose $P(k-1)$ and $P(k)$ are true for some positive integer $k$. Then \[f_2 + f_4 + \dots + f_{2k} + f_{2k+2} = f_{2k+1} - 1 + f_{2k+2} = f_{2k+3} - 1,\] whence $P(k+1)$ is true. Thus, by strong induction, the desired result holds.
\end{proof}

\begin{problem}
    Show that if $m$ and $m+2$ are both primes with $m > 3$, then $m+1$ is divisible by 6.
\end{problem}
\begin{proof}
    For primes greater than 3, the only possible residues modulo 6 are 1 and 5. This immediately gives $m \equiv 5 \pmod{6}$, whence $m + 1 \equiv 0 \pmod{6}$.
\end{proof}

\begin{problem}
    Let $a, b$ be integers, not both zero. If $\gcd{a, b} = 1$, show that the possible values for $\gcd{a+b, a-b}$ can be 1 and 2.
\end{problem}
\begin{proof}
    Let $b = a - k$ for some integer $k$. Then \[1 = \gcd{a, b} = \gcd{a, a-k} = \gcd{a, -k} = \gcd{a, k}.\] Note also that \[\gcd{a+b, a-b} = \gcd{2a, a-b} = \gcd{2a, k}.\]

    If $k$ is odd, then $\gcd{2, k} = 1$ so \[\gcd{2a, k} = \gcd{2, k} \times \gcd{a, k} = 1.\]

    If $k$ is even, then $\gcd{2, k} = 2$ so \[\gcd{2a, k} = \gcd{2, k} \times \gcd{a, k} = 2.\]
\end{proof}