\section{Self-Practice A10.4}

\begin{problem}
    If $\arg{z - 2} = 2\pi/3$ and $\abs{z} = 2$, determine $\arg{z}$.
\end{problem}

\begin{problem}
    $z$ is a complex number such that $\arg{z - 1} = \pi/3$ and $\arg{z - \i} = \pi/6$. By finding the Cartesian equations of the two half-lines, or otherwise, find the value of $\arg{z}$.
\end{problem}

\begin{problem}
    The complex number $z$ is given by $z = r\e^{\i \t}$, where $r > 0$ and $0 \leq \t \leq \pi/2$.

    \begin{enumerate}
        \item Given that $w = \bp{1 - \i \sqrt3}z$, find $\abs{w}$ in terms of $r$ and $\arg w$ in terms of $\t$.
        \item Given that $r$ has a fixed value, draw an Argand diagram to show the locus of $z$ as $\t$ varies. On the same Argand diagram, show the corresponding locus of $w$. You should identify the modulus and argument of the end-point of each locus.
    \end{enumerate}
\end{problem}

\begin{problem}
    The complex number $z$ satisfies the equation $\abs{z} = \abs{z + 2}$. Show that the real part of $z$ is $-1$. The complex number $z$ also satisfies the equation $\abs{z} = 3$. The two possible values of $z$ are represented by the points $P$ and $Q$ in an Argand diagram. Draw a sketch showing the positions of $P$ and $Q$, and calculate the two possible values of $\arg z$, giving your answers in radians correct to 3 significant figures.

    It is given that $P$ and $Q$ lie on the locus $\abs{z - a} = b$, where $a$ and $b$ are real, and $b > 0$. Give a geometrical description of this locus, and hence find the least possible value of $b$ and the corresponding value of $a$.
\end{problem}

\begin{problem}
    The complex number $z$ is given by $z = x + \i y$, where $x > 0$ and $y > 0$. Sketch an Argand diagram, with origin $O$, showing points $P$, $Q$ and $R$ representing $z$, $2\i z$ and $(z + 2\i z)$ respectively. State the size of angle $POQ$, and describe briefly the geometrical relationship between $O$, $P$, $Q$ and $R$.

    \begin{enumerate}
        \item Given that $x = 2y$, show that $R$ lies on the imaginary axis.
        \item Given that $y = 2x$, show that the point representing $z^2$ is collienar with the origin and the point $R$.
        \item Given that $\abs{z} \leq 2$ and $\arctan \frac12 \leq \arg z \leq \arctan 2$, calculate the area of the region in which the point $P$ can lie.
    \end{enumerate}
\end{problem}

\begin{problem}
    A complex number $z$ satisfies $\abs{z - a} = a$, $a \in \RR^+$.

    \begin{enumerate}
        \item The point $P$ represents the complex number $w$, where $w = 1/z$, in an Argand diagram. Show that the locus of $P$ is a straight line.
        \item Sketch both loci on the same diagram and show that the two loci do not intersect if $0 < a , 1/2$.
        \item For $a = 1/2$, find the range of values of $\arg{z - 1/a}$, giving your answer correct to $0.1\deg$. State the limit of $\arg{z - 1/a}$ when $a$ approaches 0.
    \end{enumerate}
\end{problem}

\begin{problem}
    Sketch, on an Argand diagram, the locus representing the complex number $z$ for which \[\abs{z - 4 - 4\i} = 2.\]

    \begin{enumerate}
        \item Given that $a$ is the least possible value of $\abs{z}$, find $a$.
        \item The complex number $p$ is such that \[\abs{p - 4 - 3\i} = 2 \quad \land \quad \abs{p} = a.\] State the exact value of $\arg p$.
        \item Deduce the greatest value of $\arg{z/p}$, giving your answer correct to 2 decimal places.
    \end{enumerate}
\end{problem}

\begin{problem}[\chili]
    On an Argand diagram, the point $U$ represents the complex number $z$, and the points $V$ and $W$ represent the complex numbers $z^2$ and $z^2 + 1$ respectively.

    \begin{enumerate}
        \item \begin{enumerate}
            \item Given that $\arg{z}= \a$, where $\pi/4 < \a < \pi/2$, so that $U$ lies on the half-line $L_1$ with equation $y = x \tan \a$ for $x > 0$, show that $V$ lies on the half-line $L_2$ with equation $y = x \tan 2\a$ for $x < 0$. Find the equation of the locus $L_3$ of $W$.
            \item The points $E$ and $F$ represent the values of $z$ for which $W$ coincides with $U$. Find the value of $\a$ for which the common point of $L_1$ and $L_3$ is either $E$ or $F$.
        \end{enumerate}
        \item Given instead that $\abs{z} = k$, where $k > 0$, so that $U$ lies on a circle $C$, show that $W$ lies on a circle $C'$, and find its centre and radius. Find the value of $k$ for which the common points of $C$ and $C'$ are $E$ and $F$.
    \end{enumerate}
\end{problem}