\section{Assignment B14}

\begin{problem}
    \begin{enumerate}
        \item Explain why the Euler method will fail for the initial-value problem \[\der{y}{x} = y \cos \sqrt x, \qquad y(0) = 0,\] where $y = y(x)$ satisfies that differential equation and is not a constant.
        \item Suppose the initial condition for the problem in part (a) is now $y(0) = 10$. Use the improved Euler method with a step size of $0.1$ to find, to three decimal places, an estimate for $y(0.1)$.
        \item Solve the differential equation \[\der{y}{x} = \frac12 y (2-x), \qquad y > 0, \quad y(0) = 10,\] expressing $y$ in terms of $x$, and simplifying your answer as far as possible.
        \item Explain why the solution found in part (c) will give a reasonable estimate for $y(0.1)$ in part (b).
    \end{enumerate}
\end{problem}
\begin{solution}
    \begin{ppart}
        By the Euler method, \[y_1 = y_0 + \D x (y_0 \cos \sqrt{x_0}) = 0 + \D x (0 \cos 0) = 0.\] It follows that $y_n = 0$ for all $n \in \NN$, whence $y$ is the zero function. However, because $y$ is not a constant function, $y$ cannot be the zero function, a contradiction. Hence, the Euler method fails.
    \end{ppart}
    \begin{ppart}
        Let $\D x = 0.1$, $y_0 = 10$ and $x_n = n \D x$.
        \begin{align*}
            \wt{y_1} &= y_0 + \D x (y_0 \cos \sqrt{x_0}) = 11,\\
            y_1 &= y_0 + \frac12 \D x \bs{y_0 \cos \sqrt{x_0} + \wt{y_1} \cos \sqrt{x_1}} = 11.023. \todp{3}
        \end{align*}
        Hence, $y(0.1) \approx 11.023$.
    \end{ppart}
    \begin{ppart}
        The given differential equation rewrites as \[\frac1y \der{y}{x} = \frac12 (2-x).\] Integrating both sides with respect to $x$, we obtain \[\int \frac1y \d y =\int \frac12 (2-x) \d x,\] which evaluates to \[\ln y = x - \frac14 x^2 + C_1,\] whence \[y = C\exp{x - \frac14 x^2}.\] Since $y(0) = 10$, we have $C = 10$. Thus, \[y = 10\exp{x - \frac14 x^2}.\]
    \end{ppart}
    \begin{ppart}
        For small $x$, we have that \[y \cos \sqrt{x} \approx y \bs{1 - \frac{(\sqrt x)^2}{2}} = \frac12 y (2-x),\] whence the two differential equations and thus their solutions are approximately equal. Since $x = 0.1$ is small, the solution found in part (c) will give a reasonable estimate for $y(0.1)$ in part (b).
    \end{ppart}
\end{solution}

\begin{problem}
    A particle moves along a straight line which passes through a fixed point $O$. It is acted on by two resistive forces, one of which is proportional to its displacement $x$ from $O$ while the other is proportional to its speed $v$. As a result, the motion of the particle is governed by the differential equation \[v \der{v}{x} = -7x - 24v.\]

    Given that $v = 121$ when $x = 0$, estimate the value of $v$ when $x = 1$ using
    \begin{enumerate}
        \item one iteration of Euler's method;
        \item one iteration of the improved Euler method.
    \end{enumerate}
    Hence, explain why $v$ is approximately a linear function for $0 \leq x \leq 1$.

    By considering the values of $x/v$ for $0 \leq x \leq 1$, use the given differential equation to find an expression for this linear function.
\end{problem}
\begin{solution}
    Rewriting the given differential equation, we obtain \[\der{v}{x} = -\frac{7x}{v} - 24.\] Let $f(x, v) = -7x/v - 24$ with initial value $v_0 = 121$ and $x_0 = 0$. Take a step size of $\D x = 1$, so $x_n = n \D x$.
    \begin{ppart}
        By the Euler method, $v_1 = v_0 + \D x f(x_0, v_0) = 97$. Thus, $y(1) \approx 97$.
    \end{ppart}
    \begin{ppart}
        By the improved Euler method,
        \begin{align*}
            \wt{v_1} &= v_0 + \D x f(x_0, v_0) = 97,\\
            v_1 &= v_0 + \frac12 \D x \bs{f(x_0, v_0) + f(x_1, \wt{v_1})} = 96.964.
        \end{align*}
        Thus, $y(1) \approx 96.964$.
    \end{ppart}

    The gradient of $v$ at $x = 0$ is $f(x_0, v_0) = -24$, which is very close to the gradient of $v$ at $x=1$, which is $f(x_1, \wt{v_1}) = -24.072$. Since the gradient of $v$ is approximately constant for $0 \leq x \leq 1$, we have that $v$ is approximately a linear function on that interval.

    Observe that for $0 \leq x \leq 1$, we have that $v \geq 96$, so $x/v \approx 0$. Thus, $\derx{v}{x} \approx -24$, whence $v = -24 x + C$. Since $v = 121$ when $x = 0$, we have $v \approx -24x + 121$.
\end{solution}

\begin{problem}
    The function $y = y(x)$ satisfies \[\der{y}{x} = \frac15 (\tan x + x^3 y).\] The value of $y(h)$ is to be found, where $h$ is a small positive number, and $y(0) = 0$.
    \begin{enumerate}
        \item Use one step of the improved Euler method to find an alternative approximation to $y(h)$ in terms of $h$.
        \item It can be shown that $y = y(x)$ satisfies \[y(h) = \e^{0.05h^4} \int_0^h \frac{\tan x}{5} \e^{-0.05x^4} \d x.\] Assume that $h$ is small and hence find another approximation to $y(h)$ in terms of $h$.
        \item Discuss the relative merits of these two methods employed to obtain these approximations.
    \end{enumerate}
\end{problem}
\begin{solution}
    \begin{ppart}
        Let $f(x, y) = (\tan x + x^3 y)/5$ with initial values $y_0 = 0$ and $x_0 = 0$. Take a step size of $\D x = h$. By the improved Euler method,
        \begin{align*}
            \wt{y_1} &= y_0 + \D x f(x_0, y_0) = 0,\\
            y_1 &= y_0 + \frac12 \D x \bs{f(x_0, y_0) + f(x_1, \wt{y_1})} = 0 + \frac12 h \bs{0 + \frac15(\tan h + 0)} = \frac{h \tan h}{10}.
        \end{align*}
        Hence, \[y(h) \approx \frac{h\tan h}{10}.\]
    \end{ppart}
    \begin{ppart}
        Since $h$ is small, we have that $\e^{0.05h^4} \approx 1$. Furthermore, since we are integrating over the interval $x \in [0, h]$, the integrand $(1/5) \tan x \e^{-0.05x^4}$ can likewise be approximated by $x/5$. Our integral hence transforms to \[y(h) = \e^{0.05h^4} \int_0^h \frac{\tan x}{5} \e^{-0.05x^4} \d x \approx \int_0^h \frac{x}{5} \d x = \frac{h^2}{10}.\]
    \end{ppart}
    \begin{ppart}
        The improved Euler method involves more steps, while the approximation in (b) is more direct.
    \end{ppart}
\end{solution}