\section{Self-Practice B14}

\begin{problem}
    A solution of the differential equation \[\der{x}{t} + x \cot t = \csc t\] has $x = 0$ when $t = 1$.
    \begin{enumerate}
        \item Use the improved Euler method with a step size of $0.5$ to estimate $x$ at $t = 2$.
        \item Solve the differential equation to find the exact value of $x$ when $t = 2$.
        \item By considering the gradients of the curve at $t = 1.5$, $t = 2$ and $t = 2.5$, comment
        on the accuracy of the Euler method to estimate $x$ at $t = 2$ and $t = 3$.
    \end{enumerate}
\end{problem}

\begin{problem}
    For the differential equation \[\der{y}{x} + \frac{y}{x} = 3x,\] consider the solution curve passing through the point $(1, 2)$.
    \begin{enumerate}
        \item Compute the Euler approximation to $y(1.1)$ using step-size $h = 0.1$.
        \item State, giving a reason, if you would expect the estimate in part (a) to be an underestimate or an overestimate of the true value.
        \item Find the general solution of the differential equation \[\der{y}{x} + \frac{y}{x} = 3x.\]
        \item Hence, find the solution curve that passes through the point $(1, 2)$ and calculate
        the percentage error of your estimate in (a).
    \end{enumerate}
\end{problem}

\begin{problem}
    A differential equation is of the form \[\der{y}{x} = \frac{y}{x} {f\bp{\frac{y}{x}}}.\]

    \begin{enumerate}
        \item By using the substitution $z = y/x$, show that \[\ln \abs{x} = \int \frac1{z\bs{f(z) - 1}} \d z.\]
        \item A particular solution of the differential equation \[\der{y}{x} = \frac{y}{x} \ln{\frac{x}{y}}\] has $y = 1$ when $x = 1$. Find $y$ in terms of $x$.
        \item Copy and complete the table below, using the improved Euler method with step size $0.5$ to estimate $y$ at $x = 2$.
        \begin{table}[H]
            \centering
            \begin{tabular}{|c|c|c|c|c|}
            \hline
            $x$ & $y$ & $\derx{y}{x}$ & $\wt{y}$ & $\D y / \D x$ \\ \hline\hline
            1 & 1 & 0 & 1 & 0.13516 \\ \hline
            1.5 & 1.06758 & & & \\ \hline
            2 & & \cellcolor{black!10} & \cellcolor{black!10} &  \cellcolor{black!10} \\ \hline
            \end{tabular}
        \end{table}
        \item Use a graphical method to explain whether the estimated value of $y$ found in part (c) is an under-estimate or over-estimate.
    \end{enumerate}
\end{problem}

\begin{problem}
    The variables $P$ and $t$ are related by the ``modified'' logistic equation \[\der{P}{t} = \frac1{10} P\bp{1 - \frac{P}{10}} (P - 1).\] The differential equation is used to model the size, $P$ (in thousands), of species of wolves in time, $t$ (in years) in a given habitat.

    \begin{enumerate}
        \item Biologists observe that if the population of wolves is ``too small'', adults run the risk of being unable to find a mate, resulting in a decrease to the population.
        \begin{enumerate}
            \item Explain how the model accounts for this observation.
            \item State the maximum population that the resources in the habitat can support.
            \item Find the equilibrium solutions.
            \item Sketch the possible solution curves for $P$ as a function of $t$.
        \end{enumerate}
        \item You are given that the population is now 2000.
        \begin{enumerate}
            \item Copy and complete the table showing the use of the improved Euler method with step size $0.5$ to estimate the population in a year's time.
            \begin{table}[H]
                \centering
                \begin{tabular}{|c|c|c|c|c|}
                \hline
                $t$ & $P$ & $\derx{P}{t}$ & $\wt{P}$ & $\D P / \D t$ \\ \hline\hline
                0 & 2 & 0.16 & 2.08 & 0.16896 \\ \hline
                0.5 & 2.08448 & & & \\ \hline
                1 & & \cellcolor{black!10} & \cellcolor{black!10} &  \cellcolor{black!10} \\ \hline
                \end{tabular}
            \end{table}
            \item How could the accuracy of the numerical method in part (b)(i) be improved?
        \end{enumerate}
        \item Suppose the wolves are being hunted at a fixed rate $E$ (in thousands per year). Write down the new model for the population.
    \end{enumerate}
\end{problem}