\section{Self-Practice B1}

\begin{problem}
    The equations of the curves $C_{1}$ and $C_{2}$ are given by $y=\frac{2x+1}{x-3}$ and $3(x-1)^{2}+4y^{2}=12$ respectively. Sketch $C_{1}$ and $C_{2}$ on the same diagram, stating the exact coordinates of any points of intersection with the axes and the equations of any asymptotes.
\end{problem}

\begin{problem}
    The curve $C$ has equation $y=\frac{x^{2}-4x}{x^{2}-9}$.

    \begin{enumerate}
        \item Express $y$ in the form $P+\frac{Q}{x-3}+\frac{R}{x+3}$, where $P$, $Q$ and $R$ are constants.
        \item Sketch $C$, showing clearly the asymptotes and the coordinates of the points of intersection with the axes.
    \end{enumerate}
\end{problem}

\begin{problem}
    The curve $C$ has the equation $y=\frac{x^{2}+px-q}{x+r}$. It is given that $C$ has a vertical asymptote at $x=-3$ and intersects the $x$-axis at $x=-2$ and $x=1$.

    \begin{enumerate}
        \item Determine the values of $p$, $q$ and $r$.
        \item State the equation of other asymptote(s).
        \item Prove, using an algebraic method, that $y$ cannot lie between two values which are to be determined.
        \item Hence, sketch $C$, labelling clearly the axial intercepts, asymptotes and the coordinates of any turning points.
        \item Deduce the number of roots of the equation $3x^{4}+3x^{3}-6x^{2}-x-3=0$.
    \end{enumerate}
\end{problem}

\begin{problem}
    The curve $C$ has equation $y^{2}=5x^{2}+4$.

    \begin{enumerate}
        \item Sketch $C$, indicating clearly the axial intercepts, the equations of the asymptotes and the coordinates of the stationary points.
        \item Hence by inserting a suitable graph, determine the range of values of $h$, where $h$ is a positive constant, such that the equation $5x^{2}+4=h^{2}\bp{1-x^{2}}$ has no real roots.
    \end{enumerate}
\end{problem}

\begin{problem}
    The curve $C$ has equation $y=\frac{mx^{2}+2x+m}{x}$, where $m$ is a non-zero constant.

    \begin{enumerate}
        \item Find the range of values of $m$ for $C$ to cut the $x$-axis at two distinct points.
        \item For $m=\frac{1}{2}$, find the equations of the asymptotes of $C$.
        \item Hence, sketch the curve $C$ for $m=\frac{1}{2}$: indicating clearly the asymptotes, any turning points and axial intercepts.
        \item By drawing a sketch of another suitable curve in the same diagram as your sketch of $C$, deduce the number of real roots of the equation $x^{2}+4x+1=-2x \e^{x}$.
    \end{enumerate}
\end{problem}

\begin{problem}
    The curve $C_{1}$ has equation $\frac{(x-1)^{2}}{4}=\frac{y^{2}}{9}+4.$
    
    Sketch $C_{1}$, making clear the main relevant features, and state the set of values that $x$ can take.

    Another curve $C_{2}$ is defined by the parametric equations \[x=\dfrac{2}{t^{2}+1},\quad y=3\sqrt{t}\ln t, \quad t > 1.\] Use a non-graphical method to determine the set of possible values of $x$. Sketch the curve $C_{2}$, labelling all axial intercepts and asymptotes (if any) clearly. 
    
    Hence, \textbf{without solving the equation}, state the number of real roots to the equation \[9\bp{\frac{2}{t^{2}+1}-1}^{2}=4\bp{3\sqrt{t}\ln t}^{2}+144,\] explaining your reason(s) clearly.

    Given that $k>0$, state the smallest integer value of $k$ such that the equation \[9\bp{\frac{2}{t^{2}+1}+k-1}^{2}=4\bp{3\sqrt{t}\ln t}^{2}+144,\] has exactly one real root which is positive.
\end{problem}

\begin{problem}[\chili]
    \begin{enumerate}
        \item An ellipse of equation $\frac{x^{2}}{a^{2}}+\frac{y^{2}}{b^{2}}=1$, where $0<b<a$, has two points called foci $F_1(-c, 0)$ and $F_2(c, 0)$. The definition of the ellipse is such that for every point $P$ on the ellipse, the sum of the distance of $P$ to $F_1$ and $F_2$ is always a constant $k$.
        \begin{enumerate}
            \item By considering a suitable point on the ellipse, determine the value of $k$ in terms of $a$ and/or $b$.
            \item By considering another suitable point on the ellipse, find $c$ in terms of $a$ and $b$.
        \end{enumerate}
        \item A hyperbola with equation $(y-h)^2 - 1 =\frac14(x-k)^2$ has $y=\frac12 x+\frac32$ as one of its asymptotes, and the point $(1,3)$ is on the hyperbola. Find the values of $h$ and $k$.
    \end{enumerate}
\end{problem}