\section{Self-Practice A10.2}

\begin{problem}
    The complex numbers $2\e^{\i \pi /12}$ and $2\e^{\i (5\pi/12)}$ are represented by the points $A$ and $B$ respectively in an Argand diagram with origin $O$. Show that the triangle $OAB$ is equilateral.
\end{problem}

\begin{problem}
    The complex numbers $z$ and $w$ are such that \[\abs{z} = 2, \quad \arg{z} = -\frac{2\pi}3, \qquad \tand \qquad \abs{w} = 5, \quad \arg{w} = \frac{3\pi}4.\]

    \begin{enumerate}
        \item Find the exact values of the modulus and argument of $w/z^2$. Hence, represent $z$, $w$ and $w/z^2$ clearly in an Argand diagram.
        \item Express $w/z^2$ in the exponential form. Hence, or otherwise, find the smallest positive integer $n$ such that $(w/z^2)^n$ is a real number.
    \end{enumerate}
\end{problem}

\begin{problem}
    Express $\frac{\cot \t + \i}{\cot \t - \i}$ in the exponential form.
\end{problem}

\begin{problem}
    \textbf{Do not use a calculator in answering this question.}

    Two complex numbers are $z_1 = 2\bp{\cos \frac\pi{18} - \i \sin \frac\pi{18}}$ and $z_2 = 2\i$.

    \begin{enumerate}
        \item Show that \[\frac{z_1^2}{z_1\conj} + z_2 = \sqrt3 = \i.\]
        \item A third complex number, $z_3$, is such that \[\bp{\frac{z_1^2}{z_1\conj} + z_2} z_3 \in \RR \quad \tand \quad \abs{\bp{\frac{z_1^2}{z_1\conj} + z_2} z_3} = \frac23.\] Find the possible values of $z_3$ in the form of $r\bp{\cos \t + \i \sin \t}$, where $r > 0$ and $-\pi < \t \leq \pi$.
    \end{enumerate}
\end{problem}

\begin{problem}
    \textbf{Do not use a calculator in answering this question.}

    The complex numbers $z$ and $w$ satisfy the following equations: \[w - z = 1 - \sqrt3, \qquad \i z + w = \bp{\sqrt3 + 1} \i.\] Find $w$ in the form $r\e^{\i \t}$, where $r > 0$ and $-\pi < \t \leq \pi$. Give $r$ and $\t$ in exact form.

    Hence, find the three smallest positive whole number values of $n$ for which $(iw)^n$ is an imaginary number.
\end{problem}

\begin{problem}[\chili]
    It is given that $z = \cos \t + \i \sin \t$, where $0 < \t < \pi/2$.

    \begin{enumerate}
        \item Show that $\e^{\i (\t - \pi/2)} = \sin\t - \i \cos \t$.
        \item Hence, or otherwise, show that $\arg{1 - z^2} = \t - \pi/2$ and find the modulus of $1 - z^2$.
        \item Hence, represent the complex number $1 - z^2$ on an Argand diagram.
        \item Given that $\frac{z\conj}{z^3 \bp{1 - z^2}}$ is real, find the possible values of $\t$.
    \end{enumerate}
\end{problem}