\section{Self-Practice A14B}

\begin{problem}
    A crossword puzzle is published in The Times each day of the week, except Sunday. A man is able to complete, on average, 8 out of 10 of the crossword puzzles.

    \begin{enumerate}
        \item Find the expected value and the standard deviation of the number of completed crossword puzzles in a given week.
        \item Show that the probability that he will complete at least 5 in a given week is $0.655$.
        \item Given that he completes the puzzle on Monday, find the probability that he will complete at least 4 in the rest of the week.
        \item Find the probability that, in a period of 4 weeks, he completes 4 or less in only one of the 4 weeks.
    \end{enumerate}
\end{problem}

\begin{problem}
    There is a lift on the ground floor of an old 50-storey building. This lift serves only the first 25 floors with an average number of breakdowns of 2 per week. On the 25th floor, there is another lift serving the 26th to the 50th floor with an average number of breakdowns of $0.5$ per week, independent of the other lift. Find, correct to 3 decimal places, the probability that

    \begin{enumerate}
        \item both lifts do not break down in a particular week,
        \item there are not more than 2 breakdowns altogether in a particular day,
        \item in a period of 7 days, there are 6 days on which there are not more than 2 breakdowns altogether.
    \end{enumerate}
\end{problem}

\begin{problem}
    The centre pages of the `8 days' magazine consist of 1 page of film and theatre reviews and 1 page of classified advertisements. The number of misprints in the reviews has a Poisson distribution with mean $2.3$ and the number of misprints in the classified section has a Poisson distribution with variance $1.7$.

    Find the probabilities that, on the centre pages, there will be
    \begin{enumerate}
        \item no misprints,
        \item 5 misprints.
    \end{enumerate}

    Given that there are 5 misprints in the centre pages, find the probability that 2 of them occur in the classified section.
\end{problem}

\begin{problem}
    The student resource centre has 4 rooms which can be booked for students' activities in a day at a time. Requests of the booking of a room take place independently with a mean of 4 requests per day.

    \begin{enumerate}
        \item Find the probability that not all requests for the booking of a room can be met on any particular day.
        \item On any particular day during which more than 2 requests are received for the booking of a room, find the probability that all requests for the booking of a room can be met on that day.
        \item Find the probability that there are at most 3 days in a particular week (taking 1 week to be 5 schooling days) during which not all requests for the booking of a room for that day can be met.
        \item Find the least number of rooms that the student resource centre should have so that, on any particular day, the probability that a request for the booking of a room for that day has to be refused is less than 0.05.
    \end{enumerate}
\end{problem}

\begin{problem}
    A jeweller sells rubies and diamonds. The average number of rubies and diamonds sold per week is $1.8$ and $2.7$ respectively.

    \begin{enumerate}
        \item Find the probability that exactly two rubies are sold in a given week.
        \item Find the probability that exactly 4 diamonds are sold in a given two-week period.
        \item Find the probability that the total number of jewels sold in a given week is at least 4.
        \item Given that less than 3 jewels are sold in a given week, find the probability that the number of rubies sold is more than the diamonds sold.
    \end{enumerate}
\end{problem}

\begin{problem}
    The number of printing errors, $Y$, in a page of a book follows a Poisson distribution with standard deviation $\s$. Given that $3\P{Y = 2} = 16\P{Y = 4}$, determine the value of $\s$ and hence state the mean of $Y$.

    \begin{enumerate}
        \item The probability that there are no errors in $k$ consecutive pages is less than $0.01$. Determine the least $k$ value.
        \item For a book of 100 pages, find the probability that at least one page has at least four errors.
    \end{enumerate}
\end{problem}

\begin{problem}
    In order to be offered a scholarship, a candidate has to pass two rounds of interview (it is assumed that all interviewers' decisions are independent).

    In the first round, there will be a panel of 10 interviewers and the probability of each interviewer passing a candidate is $0.9$. The candidate fails to qualify for the second round if more than one interviewer decides not to pass him or her.

    \begin{enumerate}
        \item Find the probability that a candidate passes the first round of interview.
    \end{enumerate}
    
    In the second round, there will be a panel of 5 interviewers and the probability of each interviewer passing a candidate is $0.8$. The candidate is offered a scholarship only if all interviews pass him or her in the second round.

    \begin{enumerate}
        \setcounter{enumi}{1}
        \item Show that the probability that the candidate is offered the scholarship is $0.241$, correct to three significant figures.
        \item There are $n$ candidates going for the interviews. Find the smallest $n$ such that there is at least a 98\% chance of 2 or more candidates being offered the scholarship.
    \end{enumerate}
\end{problem}

\begin{problem}
    The two most common types of disciplinary offences in a particular boy school is keeping long hair and failure to wear the school badge. The mean number of disciplinary offences recorded per day involving long hair is $1.12$. Assuming that each school week consists of five school days, the mean number of disciplinary offences recorded per school week involving failure to wear the school badge is $4.2$. The number of cases for each disciplinary offence is assumed to have an independent Poisson distribution.

    \begin{enumerate}
        \item Find the probability that at most 9 cases of disciplinary offence are recorded in a given school week.
        \item In a school week in which there are more than 7 cases of disciplinary offence involving long hair, find the probability that at most 9 cases of disciplinary offence are recorded.
        \item Calculate the probability that on a Thursday in a particular school week, it is the third day in the school week in which the discipline master caught at least 4 students having long hair in a day (you may assume that Monday is the first day of school for a school week).
        \item Explain why the Poisson distribution may not be a good model for the number of disciplinary cases involving long hair, in a school year.
    \end{enumerate}
\end{problem}

\begin{problem}
    In a small company, the employees send an average of $1.2$ print jobs to the colour print and $\a$ print jobs to the laser printer per day. It is assumed that the print jobs are independent.

    \begin{enumerate}
        \item Given that on 1 in 100 working days there are no print jobs for both printers, show that $\a=3.41$ correct to 3 significant figures.
        \item Let $E$ be the event that more than 3 print jobs were sent in on a working day. Find $\P{E}$. Hence, find the probability that the first occurrence of event $E$ happens before the 5th working day.
        \item A typical working day consists of 8 hours of work. Find the probability that more than half of the total print jobs sent during a typical working day occurs within the first hour of work, given that there was a total of 3 print jobs for the day.
    \end{enumerate}
\end{problem}

\begin{problem}
    A firm investigated the number of employees suffering injuries whilst at work. The results recorded below were obtained for a 52-week period:

    \begin{table}[H]
        \centering
        \begin{tabular}{|r|c|c|c|c|c|}
        \hline
        Number of employees injured in a week & 0 & 1 & 2 & 3 & 4 or more \\ \hline
        Number of weeks & 31 & 17 & 3 & 1 & 0 \\ \hline
        \end{tabular}
    \end{table}

    Give reasons why one might expect this distribution to approximate to a Poisson distribution. Evaluate the mean and variance of the data and explain why this gives further evidence in favour of a Poisson distribution. Using the calculated value of the mean, find the theoretical frequencies of a Poisson distribution for the number of weeks in which 0, 1, 2, 3, 4 or more employees were injured.
\end{problem}

\begin{problem}[\chili]
    The Entrepreneur Club is in charge of selling the school's tee shirt. Based on the sales record of the club, it was found that the monthly demand for tee shirt size XS has a Poisson distribution with mean 2 and the monthly demand for tee shirt size XXL has a Poisson distribution with mean 3. The club kept a monthly stock of 3 and 4 for tee shirt sizes XS and XXL respectively.

    \begin{enumerate}
        \item Calculate the probability that there is more than one XS size tee shirt being sold in a day, assuming there are 30 days in a month.
        \item Calculate the probability that the club will not meet the demand for either XS or XXL tee shirts in a month.
        \item Find the most probable number of XXL tee shirts sold in a month.
        \item Determine the least number of stock needed each month for the XS tee shirts in order to meet the demand with a probability of at least 0.95.
    \end{enumerate}
\end{problem}